### 2017-02-09 Manuscript/SW_early_life_telomeres.Rmd
Edits

###Discussion
Here we use a long-term, multi-cohort dataset to assess the relationships between spatio-temporal variation in the ecological environment and lifelong telomere dynamics in a contained population of Seychelles warblers. We found that telomere length decreases with age, and that this decrease is greatest in early life. Telomere length decreased with age in almost all of the `r length(table(dd$Cohort))` cohorts studied, but telomere length, and the rate of decrease of telomere length with age, varied substantially among cohorts. Despite an overall trend for telomere shortening in the Seychelles warbler, we found that telomere length increased within some individuals at particular life stages, and that the extent of these increases cannot be explained solely by qPCR measurement error. Finally, we found that telomeres are related to tarsus length in a sex-specific manner, and that telomere length is positively associated with temporal fluctuations in food availability.

Our study adds to the substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g. @Frenck1998; @Haussmann2003; @Heidinger2012]. However, we also found that despite an overall trend for shortening, telomere length both increased and decreased, especially after the juvenile period. Importantly, these increases were observed in longitudinal as well as cross-sectional data, suggesting that selective disappearance of individuals with shorter telomeres does not explain this pattern entirely. Longitudinal increases in measured telomere length have been observed in humans and [reviewed in @Steenstrup2013] wild animals [@Kotrschal2007; @Fairlie2016]. The most commonly invoked explanation for increases in telomere length is measurement error, which can be a particular problem in qPCR-based telomere studies [@Nussey2014; @Steenstrup2013; @Verhulst2015]. However, recent modelling work suggests that longitudinal telomere dynamics in humans are indeed consistent with apparent lenghtening, and that dismissing apparent telomere lengthening as solely measurment error is "too strong" without additional data [@Bateson2016a]. Here, we explicitly compare intra-individual variation among samples to variation among sample replicates, on a large scale. Our results suggest that qPCR measurement error alone cannot explain observed increases in RTL observed within individuals.

Increases in telomere length were not consistent over individual lifespans, but increases occured at specific, short periods, against a backdrop of overall telomere shortening. Consistent with a pattern of sporadic changes in telomere length with age, we found that within-individual telomere measurements were only weakly correlated. These findings are in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nest-box [@Boonekamp2014] conditions, in which within-individual telomere length measurements were highly consistent, and individual-level telomere shortening occurred throughout the juvenile period and into adulthood. However, the lifelong telomere dynamics found in Seychelles warblers are strikingly similar to those found in Soay sheep [@Fairlie2016], and we anticipate that a similar pattern may be found in other wild populations.

The finding that increases in telomere length may be sporadic and overlaid onto an overall pattern of shortening is an important point when assessing the occurrence of telomere lengthening. Previously described approaches to distinguish telomere elongation from measurement error, based on assumptions about follow-up time between measurements [@Steenstrup2013], or based on measuring variance among measurements [@Simons2014], assume that telomere elongation within individuals is consistent over time. Such sporadic changes in telomere length over lifespans could occur due to changes in the composition of cell types within individual samples, or due to the actual elongation of telomeres. Determining the mechanism of these changes is essential for how we view telomeres as biomarkers of costs. For example, if telomeres can be lengthened in response to improvements in environmental conditions, this would suggest that they reflect short- to medium-term costs, rather than the cumulative costs that an individual has faced over its lifespan [@Bateson2016]. New statistical and technical approaches are therefore now required to determine the mechanisms behind increases in telomere length within individuals, so that biologically informed hypotheses about the ecological causes and consequences of these increases can be generated and tested.

A few studies have shown that temporal variation in telomere dynamics occurs in natural populations, although these have been limited in the number of seasons [@Mizutani2013; @Watson2015; @Fairlie2016]. The long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods. Our data suggest that birth year is a highly important factor in shaping telomere dynamics, but also that age-related declines in telomere length vary among cohorts. This is consistent with telomere length being controlled by both genetic and environmental factors - something that is becoming apparent from quantitative genetic studies of telomere dynamics. Moreover, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process. Research of telomere dynamics within and across multiple cohorts and populations will enable us to better understand how how and why population-level telomere dynamics vary over space and time.

We found that temporal variation in population-level food availability was positively related to telomere length. This is consistent with the strong cohort effects we found, and suggests that temporal variation in conditions may be a key driver of costs in the Seychelles warbler. Although the environmental conditions on Cousin Island are relatively benign in comparison to other island systems [e.g. @Coulson2001], substantial annual variation in rainfall does occur, with associated changes in insect abundance, and it appears that this confers a cost to Seychelles warblers. We also found evidence for sex-specific telomere dynamics, with males having longer-telomeres than females, and that this sex-difference interacts with tarsus length. If the sex-dependent relationship between telomere and tarsus lengths was due to differential growth alone we would expect the opposite pattern to that observed, as male Seychelles warblers are larger than females (Fig. 3B). One possibility is that the environment imposes differential costs on males and females: a recent study in captive zebra finches found that manipulation of dietary nutrients had sex-dependent effects on telomere dynamics [@Noguera2015]. Also worth noting is that the effect of telomere length on survival in strongest in male Seychelles warblers [@Barrett2013a], although the nature of the relationship between sex, telomeres and survival is not yet clear [@Barrett2011]

Although we found clear associations between the environment and telomere dynamics, we should bear in mind that our social and ecological variables explained a only a small proportion of the variance in RTL. A poor social and ecological environment is known to be detrimental to Seychelles warblers, both in terms of oxidative stress and survival [@VandeCrommenacker2011a; @Brouwer2012], and it is therefore perhaps surprising that these variables do not explain more variance in RTL. While measurement error clearly accounts for some of this lack of explanatory power, it is also likely that early-life RTL in the Seychelles warbler is explained by a set of environmental and genetic variables not considered here [e.g. @Bebbington2016]. A key question to be addressed is the extent to which RTL, especially in early life, reflects inheritance and parental effects [e.g. @Becker2015; @Asghar2014; @Heidinger2016]. Long-term ecological study systems are uniquely suited to addressing such questions in natural systems [@Clutton-Brock2010]. To gain a full understanding of telomere dynamics in natural systems, long-term studies combining ecological and genetic data will be required from a range of species.

>>>>

###Discussion
Here we use a long-term, multi-cohort dataset to assess lifelong telomere dynamics and the relationship between these and spatio-temporal variation in the ecological environment in a contained population of Seychelles warblers. We found that telomere length decreases with age, and that this decrease is greatest very early in life. Telomere length decreased with age in almost all of the `r length(table(dd$Cohort))` cohorts studied, but the rate of decrease of telomere length with age, and consequently the telomere length of adults, varied substantially among cohorts. Despite an overall pattern of telomere shortening with age in the Seychelles warbler, we found that our measure of telomere length increased within some individuals at particular life stages, and that the extent of these increases could not be explained solely by qPCR measurement error. Finally, we found that telomeres are related to tarsus length in a sex-specific manner, and that telomere length is positively associated with temporal fluctuations in food availability.

Our study adds to the substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g. @Frenck1998; @Haussmann2003; @Heidinger2012]. However, we also found that despite an overall trend for shortening, telomere length both increased and decreased, especially after the juvenile period. Importantly, these increases were observed in longitudinal as well as cross-sectional data, indicating that selective disappearance of individuals with shorter telomeres does not explain this pattern. Longitudinal increases in measured telomere length have been observed in humans and wild animals  [@Steenstrup2013;@Kotrschal2007; @Fairlie2016; @Hoelzl2016; @Hoelzl2016a]. The most commonly invoked explanation for increases in telomere length is measurement error, which can be a particular problem in qPCR-based telomere studies [@Nussey2014; @Steenstrup2013; @Verhulst2015]. However, recent modelling work suggests that longitudinal telomere dynamics in humans are indeed consistent with apparent lenghtening, and that dismissing apparent telomere lengthening as solely measurment error is "too strong" without additional data [@Bateson2016a]. Here, we explicitly compare intra-individual variation among samples to variation among sample replicates, on a large scale. Our results suggest that qPCR measurement error alone cannot explain observed increases in RTL observed within individuals.

Increases in telomere length were not consistent over individual lifespans, but occurred during specific, periods, against a backdrop of overall lifelong telomere shortening. This is consistent with recent findings in edible dormice *Glis glis*, in which telomere elongation was observed only later in life [@Hoelzl2016a]. Consistent with a pattern of sporadic changes in telomere length with age, we found that within-individual telomere measurements were only weakly correlated. These findings are in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nest-box [@Boonekamp2014] conditions, in which within-individual telomere length measurements were highly consistent, and individual-level telomere shortening occurred throughout the juvenile period and into adulthood. However, the lifelong telomere dynamics found in Seychelles warblers are strikingly similar to those found in Soay sheep [@Fairlie2016], and we anticipate that a similar pattern may be found in other wild populations.

The finding that increases in telomere length may be sporadic and overlaid on an overall pattern of shortening with age is an important point when assessing the occurrence of telomere lengthening. Previously described approaches to distinguish telomere elongation from measurement error, based on assumptions about follow-up time between measurements [@Steenstrup2013], or based on measuring variance among measurements [@Simons2014], assume that telomere elongation within individuals is consistent over time. Our data, and that of others [@Fairlie2016;@Hoelzl2016; @Hoelzl2016a] suggest that this is not the case.  Such inconsistent changes in telomere length over lifespans could occur due to changes in the composition of cell types within individual samples, or due to the actual elongation of telomeres [@Blackburn1989]. Determining the mechanism of these changes is essential for how we view telomeres as biomarkers of costs. For example, if telomeres can be lengthened in response to improvements in environmental conditions, this would suggest that they reflect short- to medium-term costs, rather than the cumulative costs that an individual has faced over its lifespan [@Bateson2016]. New statistical and technical approaches are therefore now required to determine the mechanisms behind increases in telomere length within individuals, so that biologically informed hypotheses about the ecological causes and consequences of these increases can be generated and tested.

Measurement of cohorts across seasons or years is required if we are to understand how the environment impacts telomere dynamics. Although a few studies have shown that temporal variation in telomere dynamics occurs in natural populations, these have been limited in the number of seasons they cover [@Mizutani2013; @Watson2015; @Fairlie2016]. TThe long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods. Our data suggest that conditions during the hatch year are a very important factor in shaping telomere dynamics, and that age-related declines in telomere length vary among cohorts. The rate of early-life telomere loss for a cohort was highly variable and had a persistent effect of the mean adult telomere length of the cohort. This effect is consistent with telomere length being strongly influenced by environmental, as well as genetic, factors - something that is becoming apparent from the, as yet, limited number of quantitative genetic studies of telomere dynamics in wild populations [e.g. @Olsson2011; @Atema2015; @Becker2015]. Moreover, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process. Research of telomere dynamics within and across multiple cohorts and populations will enable us to better understand how and why population-level telomere dynamics vary over space and time.

We found that temporal variation in insect prey availability was positively related to telomere length. This is consistent with the strong cohort effects we found, and suggests that temporal variation in environmental conditions may be a key driver of costs in the Seychelles warbler.  Although the environmental conditions on Cousin Island are relatively benign in comparison to other island systems [e.g. @Coulson2001], substantial annual variation in rainfall does occur, with associated changes in insect abundance [@Komdeur1996], and it appears that this confers a cost - in terms of intrinsic biological condition - to Seychelles warblers. Our results concur with other studies which show that early life  conditions / food availability can have a very significant and long term impacts on telomere length (and intrinsic biological condition) in captive and wild animals [e.g. @Nettle2015; @Heidinger2012; @Watson2015].

We also found evidence for sex-specific telomere dynamics: males had longer-telomeres than females. Interestingly this sex-difference interacts with tarsus length: telomere length was negatively correlated with tarsus length in both sexes, but this effect was stronger in females than males. If the sex-dependent relationship between telomere and tarsus length was due to differential growth alone we would expect the opposite pattern to that observed, as male Seychelles warblers are larger than females (Fig. 3B). One possibility is that the environment imposes differential costs on males and females: a recent study in captive zebra finches found that manipulation of dietary nutrients had sex-dependent effects on telomere dynamics [@Noguera2015]. Also worth noting is that the effect of telomere length on survival in strongest in male Seychelles warblers [@Barrett2013a], although the nature of the relationship between sex, telomeres and survival is not yet clear [@Barrett2011]

Although we found clear associations between the environment and telomere dynamics, we should bear in mind that the social and ecological variables we tested here explained only a small proportion of the variance in RTL. A poor social and ecological environment is known to be detrimental to Seychelles warblers, both in terms of oxidative stress and survival [@VandeCrommenacker2011a; @Brouwer2012], and it is therefore perhaps surprising that these variables do not explain more variance in RTL. While telomere measurement error will account for some of this lack of explanatory power, it is also likely that early-life RTL in the Seychelles warbler is explained by a set of environmental and genetic variables not considered here [e.g. @Bebbington2016]. A key question to be addressed is the extent to which RTL, especially in early life, reflects inheritance and parental effects [e.g. @Becker2015; @Asghar2014; @Heidinger2016]. For example, parental age and quality may key variables that impact the telomere dynamics of offspring in the Seychelles warbler, and will be addressed in future studies.  Long-term ecological study systems are uniquely suited to addressing such questions in natural systems [@Clutton-Brock2010]. To gain a full understanding of telomere dynamics in natural systems, long-term studies combining ecological and genetic data will be required from a range of species.

### 2017-02-09 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Edits

**Figure 1** Telomere dynamics in relation to age in Seychelles warbler cohorts. **A** RTL and age across all individuals. Points and  thin grey lines represent individual samples and birds, respectively. The thick line and shaded area represent the fitted values and 95% confidence limits of a linear regression of RTL and log-transformed age.**B** Boxplot of variation in RTL among cohorts. **C** RTL and age among cohorts. Lines represent fitted values from a linear regression and log-transformed age, and colours correspond to **B**. **D** RTL in relation to and $\Delta$Log age (i.e. within indiviual variation in log age).

**Figure 2** Longitudinal telomere dynamics in the Seychelles warbler. **A** Variation in RTL within individuals sampled at different time points. The dotted line represents parity, and thus points above and below the line represent increases and decreases in RTL, respectively. **B** Scaled density plots of repeated RTL measurements among individual samples (grey), and among different samples taken from the same individual (yellow). Areas of the density plot to the left of the dotted line represent decreases in RTL, while areas to the right represent increases.

@@

Fig2B <- ggplot(ddL,aes(col=Group,fill = Group,y = ..scaled..,x = DeltaRTL))+
    geom_density(alpha = 0.2)+
  theme_classic()+
  ylab ('Scaled density')+
  xlab(expression(Delta*'RTL'))+
    theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = 'none',
        legend.title = element_blank(),
        axis.line.x = element_line(colour = 'black'),
        axis.line.y = element_line(colour = 'black')) +
  scale_colour_manual(values = c('darkgrey','gold'))+
  scale_fill_manual(values = c('darkgrey','gold'))+
  geom_vline(xintercept = 0,lty = 2) +
  annotate('text',x = -1,y = 1,label = 'B')

Fig2C <- ggplot(temp, aes(x = Agemonths,
           y = DeltaRTL)) +
  ylab(expression(Delta*'RTL'))+
  xlab("Age (years)") +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = c('none'),
        legend.title = element_text(size = 12),
        axis.line.x = element_line(colour = 'black'),
        axis.line.y = element_line(colour = 'black')) +
  geom_point(col = "grey")+
  geom_smooth(col = 'black',fill = 'black',method = "lm",formula = y~log(x))+
  geom_abline(slope = 0,intercept = 0,lty = 2)+
  annotate('text',x = -0.2,y = 1,label = 'C')

@@

Fig3B <- ggplot(dd,
       aes(x = Tarsus,
           y = RTL,
           col = Sex)) +
  geom_point()+
  ylab('RTL') + 
  xlab('Tarsus length (mm)') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
                axis.line.x = element_line(colour = 'black'),
        axis.line.y = element_line(colour = 'black'),
        legend.position = 'none') +
  stat_smooth(method = 'lm') +
  scale_color_manual(values = c('gold2','red'))+
  annotate('text',x = 17,y = 1.7,label = 'B')

>>>>

**Figure 1** Telomere dynamics in relation to age in Seychelles warbler cohorts. **A** RTL and age across all individuals. Points and  connecting thin grey lines represent individual samples and birds, respectively. The thick line and shaded area represent the fitted values and 95% confidence limits of a linear regression of RTL and log-transformed age. **B** Boxplot of variation in RTL among cohorts. **C** RTL and age among cohorts. Lines represent fitted values from a linear regression of RTL and log-transformed age, and colours correspond to **B**. **D** RTL in relation to and $\Delta$Log age (i.e. within indiviual variation in log age).

**Figure 2** Longitudinal telomere dynamics in the Seychelles warbler. **A** Variation in RTL within individuals sampled at different time points. The dotted line represents parity, and thus points above and below the line represent increases and decreases in RTL, respectively. **B** Scaled density plots of repeated RTL measurements among individual samples, and among different samples taken from the same individual. Areas of the density plot to the left of the dotted line represent decreases in RTL, while areas to the right represent increases. **C** $\Delta$RTL in relation to age in pairs of samples taken within two years. Black line and shaded area represent fitted values and 05% confidence limits from a linear regression of RTL and log-transformed age. **D** Telomere lengthening and shortening in relation to age. Points at zero and one represent pairs of samples where RTL has decreased and increased, respectively, with point size scaled by the number of overlapping values. The grey line represents the proportion of samples in which i ncreases in RTL where observed at each age.

@@

Fig2B <- ggplot(ddL,aes(col=Group,fill = Group,y = ..scaled..,x = DeltaRTL))+
    geom_density(alpha = 0.2)+
  theme_classic()+
  ylab ('Scaled density')+
  xlab(expression(Delta*'RTL'))+
    theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = "top",
        legend.title = element_blank(),
        axis.line.x = element_line(colour = 'black'),
        axis.line.y = element_line(colour = 'black')) +
  scale_colour_manual(values = c('darkgrey','gold'))+
  scale_fill_manual(values = c('darkgrey','gold'))+
  geom_vline(xintercept = 0,lty = 2) +
  annotate('text',x = -1,y = 1,label = 'B')

Fig2C <- ggplot(temp, aes(x = Agemonths,
           y = DeltaRTL)) +
  ylab(expression(Delta*'RTL'))+
  xlab("Age (years)") +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        axis.line.x = element_line(colour = 'black'),
        axis.line.y = element_line(colour = 'black')) +
  geom_point(col = "grey")+
  geom_smooth(col = 'black',fill = 'black',method = "lm",formula = y~log(x))+
  geom_abline(slope = 0,intercept = 0,lty = 2)+
  annotate('text',x = -0.2,y = 1,label = 'C')

@@

Fig3B <- ggplot(dd,
       aes(x = Tarsus,
           y = RTL,
           col = Sex)) +
  geom_point()+
  ylab('RTL') + 
  xlab('Tarsus length (mm)') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
                axis.line.x = element_line(colour = 'black'),
        axis.line.y = element_line(colour = 'black'),
        legend.position = 'top',
        legend.title = element_blank()) +
  stat_smooth(method = 'lm') +
  scale_color_manual(values = c('gold2','red'))+
  annotate('text',x = 17,y = 1.7,label = 'B')
